---
title: "ts and tsible"
author: "sample analysis"
date: "`r Sys.Data()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(forecast) #functions for time series forecasting
library(tseries)  #statistical tests and time series analysis tools

library(tidyverse)
library(here)
library(data.table)
library(lubridate)
library(dplyr)
library(tsibble) #yearweek
library(fable) #univariate and multivariate time series forecasting models, tsibble
```

[Source](https://www.kdnuggets.com/how-to-conduct-time-series-analysis-in-r)

```{r cars}
df_raw <- fread(here("1_data","clean","df_ili.csv"))
# head(df_raw)
#shift + arrow
#ctrl + enter to execute the highlited section 
df <- df_raw %>% filter(REGION == "Alabama") %>% 
  select(c("YEAR","WEEK","TOTAL PATIENTS"))

df$TEMP <- paste0(df$YEAR,"W",df$WEEK)
df <- df %>% mutate(week = yearweek(TEMP)) %>% select(-c("TEMP"))

ds <- df %>% as_tsibble(
  index = week,
  regular = TRUE
)

ds %>% ggplot(aes(x=week, y = `TOTAL PATIENTS`)) + geom_line()

ts_data <- ts(ds$`TOTAL PATIENTS`)
class(ds)
class(ts_data)
```

```{r}
plot(ts_data)

fit <- auto.arima(ts_data)
```

## Autocorrelation Function (ACF)
 
- The Autocorrelation Function (ACF) measures how a time series is correlated with its past values. It helps identify patterns and lags in the data. 

- It shows these correlations at different time lags. 

- The ACF plot helps determine the Moving Average (MA) order ('q').

```{r}
acf(ts_data)
```

## Partial Autocorrelation Function (PACF)
 

- The Partial Autocorrelation Function (PACF) measures the correlation of a time series with its past values. - It excludes the effects of intervening lags. 
  - It helps identify the strength of direct relationships at different lags. 
  - The PACF plot displays these correlations for various time lags. 
  - The PACF plot helps identify the Auto-Regressive (AR) order ('p').

```{r}
pacf(ts_data)

```

## Ljung-Box Test
 

$$H_0: \text{The residuals are independently distributed.}$$

$$H_A: \text{The residuals are not independently distributed; they exhibit serial correlation.}$$

- The Ljung-Box test checks for autocorrelation in the residuals of a time series model. It tests if the residuals are random. 

- It tests for autocorrelation at multiple lags. 
  - A low p-value suggests significant autocorrelation. This means the model might not be a good fit.

```{r}
Box.test(fit$residuals, lag = 20, type = "Ljung-Box")
```

## Residual Analysis
 
- Residual analysis examines the differences between the observed and predicted values from a time series model. 

- It helps check if the model fits the data well.

```{r}
plot (fit$residuals, main="Residuals of ARIMA Model", ylab="Residuals")
abline(h=0, col="red")
```

## Forecasting
 
- Forecasting involves predicting future values based on historical data. Use the 'forecast' to generate these predictions.

```{r}
forecast_result <- forecast(fit) 

autoplot(forecast_result)
```
# using Fable

```{r}
fit <- ds %>%
  model(
    arima = ARIMA(`TOTAL PATIENTS`)
  )

fit %>% glance()

```

