---
title: "tsibble"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
      smooth_scroll: false
      number_sections: true
      fig_caption: TRUE
    css: "style.css"
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(fpp3)
library(tsibble) #yearweek
#univariate and multivariate time series forecasting models, tsibble
library(fable) 

library(tidyverse)
library(here)
library(data.table)
library(lubridate)
library(dplyr)
```

Reading: CH1 to CH2.5

|Type | why?|
|:----|:-----|
|Short-term forecasts| needed for the scheduling of personnel, production and transportation. As part of the scheduling process, forecasts of demand are often also required.
|Medium-term forecasts |are needed to determine future resource requirements, in order to purchase raw materials, hire personnel, or buy machinery and equipment.|
|Long-term forecasts|used in strategic planning. Such decisions must take account of market opportunities, environmental factors and internal resources.|

# Data

- must have `index`

```{r}
y <- tsibble(
  Year = 2015:2019,
  Observation = c(123, 39, 78, 52, 110),
  index = Year
)

y %>% autoplot()
```

## Olympic Running

```{r}
olympic_running %>% head()

olympic_running |> distinct(Sex)
```

## PBS

- This contains monthly data on Medicare Australia prescription data from July 1991 to June 2008. 
- These are classified according to various `concession types`, and `Anatomical Therapeutic Chemical` (ATC) indexes. 
- For this example, we are interested in the `Cost` time series (total cost of scripts in Australian dollars)
```{r}
PBS

glimpse(PBS)
```

### filter()

```{r}
PBS %>% 
  filter(ATC2 == "A10") %>% head()
```

### select()

```{r}
PBS %>% 
  filter(ATC2 == "A10") %>% 
  select(Month, Concession, Type, Cost) %>% head()
```

### summarise()

```{r}
PBS %>% 
  filter(ATC2 == "A10") %>% 
  select(Month, Concession, Type, Cost) %>% 
  summarise(TotalC = sum(Cost))
```

### mutate()

```{r}
PBS %>% 
  filter(ATC2 == "A10") %>% 
  select(Month, Concession, Type, Cost) %>% 
  summarise(TotalC = sum(Cost)) %>% 
  mutate(Cost = TotalC/1e6)

#save this for later     
a10 <- PBS %>% 
  filter(ATC2 == "A10") %>% 
  select(Month, Concession, Type, Cost) %>% 
  summarise(TotalC = sum(Cost)) %>% 
  mutate(Cost = TotalC/1e6) 
  
head(a10)  
  
```

## PRISON

```{r}
prison <- readr::read_csv("https://OTexts.com/fpp3/extrafiles/prison_population.csv")

head(prison)
unique(prison$Date)

prison %>% 
  mutate(Quarter = yearquarter(Date)) %>% head()

prison %>% 
  mutate(Quarter = yearquarter(Date)) %>% 
  select(-Date) %>% head()
```

- identify which column contains 
  - the `time index`, 
  - and which columns are `keys`. 
  - The remaining columns are `values` — there can be many value columns

```{r}
#key and index
prision <- prison %>% 
  mutate(Quarter = yearquarter(Date)) %>% 
  select(-Date) %>% 
  as_tsibble(key = c(State, Gender, Legal, Indigenous),
             index = Quarter)
```

```{r}
unique(prison$State)
```

- This tsibble contains `64` separate time series corresponding to the combinations of the 8 states, 2 genders, 2 legal statuses and 2 indigenous statuses. 

- Each of these series is 48 observations in length, from 2005 Q1 to 2016 Q4.


## ANSETT

```{r}
melsyd_economy <- ansett %>% 
  filter(Airports == "MEL-SYD", Class == "Economy") %>% 
  mutate(Passengers = Passengers/1000)


autoplot(melsyd_economy, Passengers) +
  labs(title = "Ansett airlines economy class",
       subtitle = "Melbourne-Sydney",
       y = "Passengers ('000)")
```

# Pattern 

## Trend

- A trend exists when there is a long-term increase or decrease in the data. 
- It does not have to be linear. 
  - Sometimes we will refer to a trend as `changing direction`, when it might go from an increasing trend to a decreasing trend. 
  - There is a trend in the antidiabetic drug sales data shown in Figure 2.2.

```{r}
autoplot(a10, Cost) +
  labs(y = "$ (millions)",
       title = "Australian antidiabetic drug sales")
```


## Seasonal

- A seasonal pattern occurs when a time series is affected by seasonal factors such as the `time of the year` or `the day of the week`. 

- Seasonality is always of a `fixed` and `known period`. 
- The monthly sales of antidiabetic drugs (see figure above) shows seasonality which is induced partly by the change in the cost of the drugs at the end of the calendar year. (Note that one series can have more than one seasonal pattern.)

### a10

```{r}
a10 |>
  gg_season(Cost, labels = "both") +
  labs(y = "$ (millions)",
       title = "Seasonal plot: Antidiabetic drug sales")
```

- `gg_season()` is useful in identifying years in which the pattern changes.
- There is a large jump in sales in January each year. 
  - These are probably sales in late December as customers stockpile before the end of the calendar year, but the sales are not registered with the government until a week or two later. 
  - The graph also shows that there was an unusually small number of sales in March 2008 (most other years show an increase between February and March). 
  - The small number of sales in June 2008 is probably due to incomplete counting of sales at the time the data were collected.

```{r}
a10 %>% gg_subseries(Cost) +
  labs(
    y = "$ (millions)",
    title = "Australian antidiabetic drug sales"
  )
```

### vic_elec

```{r fig.height=11}
p1 <- vic_elec |> gg_season(Demand) +
  theme(legend.position = "none") +
  labs(y="MWh", title="Electricity demand: Victoria(YEAR)")

p2 <- vic_elec |> gg_season(Demand, period = "month") +
  theme(legend.position = "none") +
  labs(y="MWh", title="Electricity demand: Victoria (MONTH)")


p3 <- vic_elec |> gg_season(Demand, period = "week") +
  theme(legend.position = "none") +
  labs(y="MWh", title="Electricity demand: Victoria (WEEK)")

p4 <- vic_elec |> gg_season(Demand, period = "day") +
  theme(legend.position = "none") +
  labs(y="MWh", title="Electricity demand: Victoria (DAY)")

p1/p2/p3/p4
```



### tourism

- `tourism` is tsibble

```{r}
tourism

holidays <- tourism %>% 
  filter(Purpose == "Holiday") %>% 
  group_by(State) %>% 
  summarise(Trips = sum(Trips))
```

- Notice what happens to `key`

```{r}
holidays

holidays %>% autoplot(Trips) +
  labs(y = "Overnight trips ('000)",
       title = "Australian domestic holidays")

```

```{r}
gg_season(holidays, Trips) +
  labs(y = "Overnight trips ('000)",
       title = "Australian domestic holidays")
```

```{r}
holidays |>
  gg_subseries(Trips) +
  labs(y = "Overnight trips ('000)",
       title = "Australian domestic holidays")
```


## Cyclic

- A cycle occurs when the data exhibit rises and falls that are not of a fixed frequency. 

- These fluctuations are usually due to economic conditions, and are often related to the `business cycle`. 
- The duration of these fluctuations is usually at least 2 years.


If the fluctuations are not of a fixed frequency then they are cyclic; if the frequency is unchanging and associated with some aspect of the calendar, then the pattern is seasonal. 

In general, the average length of cycles is longer than the length of a seasonal pattern, and the magnitudes of cycles tend to be more variable than the magnitudes of seasonal patterns.

Many time series include trend, cycles and seasonality. When choosing a forecasting method, we will first need to identify the time series patterns in the data, and then choose a method that is able to capture the patterns properly.

- When we decompose a time series into components, we usually combine the `trend` and `cycl`e into a single `trend-cycle component` (often just called the trend for simplicity). 
  - Thus we can think of a time series as comprising three components: a `trend-cycle component`, a `seasonal component`, and a remainder component (containing anything else in the time series). 
  - For some time series (e.g., those that are observed at least daily), there can be more than one seasonal component, corresponding to the different seasonal periods.