---
title: "correlation"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
      smooth_scroll: false
      number_sections: true
      fig_caption: TRUE
    css: "style.css"
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(fpp3)
library(tsibble) #yearweek
#univariate and multivariate time series forecasting models, tsibble
library(fable) 

library(tidyverse)
library(here)
library(data.table)
library(lubridate)
library(dplyr)

#new library
library(GGally)

library(here)
source(here("source","get_fpp3.R"))
```

- Reading: CH2.6 to 2.10

|Type | why?|
|:----|:-----|
|Short-term forecasts| needed for the scheduling of personnel, production and transportation. As part of the scheduling process, forecasts of demand are often also required.
|Medium-term forecasts |are needed to determine future resource requirements, in order to purchase raw materials, hire personnel, or buy machinery and equipment.|
|Long-term forecasts|used in strategic planning. Such decisions must take account of market opportunities, environmental factors and internal resources.|

# Electricity

- The temperatures are for `Melbourne`, the largest city in Victoria, while the demand values are for the `entire state`

```{r}
vic_elec

vic_elec %>% 
  filter(year(Time) == 2014) %>% 
  autoplot(Demand) +
  labs(y = "GW",
       title = "Half-hourly electricity demand: Victoria")
```

```{r}
vic_elec |>
  filter(year(Time) == 2014) |>
  autoplot(Temperature) +
  labs(
    y = "Degrees Celsius",
    title = "Half-hourly temperatures: Melbourne, Australia"
  )
```

We can study the relationship between demand and temperature by plotting one series against the other.

```{r}
vic_elec %>% 
  filter(year(Time) == 2014) %>% 
  ggplot(aes(x = Temperature, y = Demand)) +
  geom_point() +
  labs(title="Electricity demand versus Temperature",
       x = "Temperature (degrees Celsius)",
       y = "Electricity demand (GW)")
```

It is clear that high demand occurs when temperatures are high due to the effect of air-conditioning. But there is also a heating effect, where demand increases for very low temperatures.

# Correlation 

- Sample correlation. 

$$r = \frac{\sum(x_i-\bar{x})(y_i-\bar{y})}{\sqrt(\sum(x_i-\bar{x})^2(y_i-\bar{y})^2)}$$

```{r}
visitors |>
  ggplot(aes(x = Quarter, y = Trips)) +
  geom_line() +
  facet_grid(vars(State), scales = "free_y") +
  labs(title = "Australian domestic tourism",
       y= "Overnight trips ('000)")
```

## Are they correlated to each other?

```{r}
visitors |>
  pivot_wider(values_from=Trips, names_from=State) |>
  GGally::ggpairs(columns = 2:9)
```

- In this example, mostly positive relationships are revealed, with the strongest relationships being between the neighbouring states located in the south and south east coast of Australia, namely, `New South Wales`, `Victoria` and `South Australia`. 

- Some negative relationships are also revealed between the `Northern Territory` and other regions.    - The Northern Territory is located in the north of Australia famous for its outback desert landscapes visited mostly in winter. 
  - Hence, the peak visitation in the Northern Territory is in the `July (winter) quarter` in contrast to `January (summer) quarter` for the rest of the regions.
  
## Lag

- Here the colours indicate the quarter of the variable on the vertical axis. 
- The relationship is strongly positive at lags 4 and 8, reflecting the strong seasonality in the data. The negative relationship seen for lags 2 and 6 occurs because peaks (in Q4) are plotted against troughs (in Q2)

  
```{r}
recent_production

recent_production |>
  gg_lag(Beer, geom = "point") +
  labs(x = "lag(Beer, k)")
```
  
# Autocorrelation 

autocorrelation measures the linear relationship between lagged values of a time series.

```{r}
recent_production |> ACF(Beer, lag_max = 9)

```

```{r}
recent_production |>
  ACF(Beer) |>
  autoplot() + labs(title="Australian beer production")
```

- $r_4$ is higher than for the other lags. 
  - This is due to the seasonal pattern in the data: the peaks tend to be four quarters apart and the troughs tend to be four quarters apart.

- $r_2$ is more negative than for the other lags because troughs tend to be two quarters behind peaks.

- The dashed blue lines indicate whether the correlations are significantly different from zero

## Trend and Seasonality

- When data have a `trend`, the autocorrelations for small lags tend to be large and positive because observations nearby in time are also nearby in value. 
  - So the ACF of a trended time series tends to have positive values that `slowly decrease as the lags increase`.

- When data are `seasonal`, the autocorrelations will be larger for the seasonal lags (at multiples of the seasonal period) than for other lags.

- When data are both `trended` and `seasonal`, you see a combination of these effects. The a10 data plotted (see below) shows both trend and seasonality. 
  - The slow decrease in the ACF as the lags increase is due to the trend, while the `scalloped` shape is due to the seasonality.
  
```{r}
a10 %>% 
  ACF(Cost, lag_max = 48) %>% 
  autoplot() +
  labs(title="Australian antidiabetic drug sales")
```

# White noise (see CH 2.9)

- Time series that show no autocorrelation are called `white noise`.

```{r}
set.seed(30)
y <- tsibble(sample = 1:50, wn = rnorm(50), index = sample)
y %>%  autoplot(wn) + labs(title = "White noise", y = "")

y %>% 
  ACF(wn) %>% 
  autoplot() + labs(title = "White noise")
```

- work on exercise 2.10

